---
title: "Data Visualization With R"
author: "Zack Treisman"
date: "Spring 2021"
output: 
  beamer_presentation: 
    keep_tex: yes
bibliography: ../bibliography.bib    
---

```{r setup, include=FALSE}
knitr::opts_knit$set(global.par = TRUE)

library(emdbook)
library(MASS)
library(tidyverse)
library(gridExtra)
library(lubridate)

```

```{r,include=FALSE}
par(mar = c(4, 4, 0.75, 0.5)) # Set margins
``` 


## Philosophy

Good visualizations are useful at every stage of the data analysis process from exploration to publication.

R has multiple graphics systems. We will use two:

* Base R graphics are often intuitive, but limited.
* `ggplot2` is robust and widely used. It takes some acclimatization.

Many older resources use `lattice` graphics, which started dropping in popularity as `ggplot2` took over.

Be wary of inference based on purely exploratory data analysis. If you look at your data until you find a pattern, and then test for that pattern, the significance levels of that test are inflated.

## Exploration

Data are in R, now what? 

* Check numerical summaries.

\scriptsize
```{r}
summary(ReedfrogPred) # data in emdbook
```

\normalsize
* Make some graphics!
    * Are there patterns that you expected to see?
    * Or didn't expect to see?
    * Are there problems with the data?


## Standard routines - one variable

Graphing the distribution of a single variable means representing how often it takes each possible value.

* Barplots for categorical variables. Same information as a table.
* Histograms for numeric variables. Can add a density estimate.

\scriptsize
```{r, fig.width=9, fig.height=3}
par(mfrow=c(1,2)) # show multiple base R plots at once
barplot(table(ReedfrogPred$pred))
hist(ReedfrogPred$propsurv,freq=F); lines(density(ReedfrogPred$propsurv))
```

## Standard routines - two numeric variables

Scatterplots show codistribution.

* Put the response variable on the $y$ axis.

\scriptsize
```{r, fig.width=8, fig.height=3.5}
p <- ggplot(Lily_sum, aes(vegetative, flowers)) + # data are in emdbook
  geom_point()
p
```

## Log scales

Sometimes the data look better with axes on log scales.

* Counts
* Dimesnional data

Recall log(0) is undefined so 0 values will produce warnings or errors.

\scriptsize
```{r, fig.width=8, fig.height=3.5, warning=FALSE}
p + scale_x_log10() + 
  scale_y_log10()
```

## Additional aesthetics 

Map additional variables to color, size or shape (plotting symbol).

* Shape can only be a categorical variable.
* Size can only be a numerical variable.

Can also superimpose trendlines or other model graphs.

\scriptsize
```{r, fig.width=8, fig.height=3.5, message=FALSE}
ggplot(Lily_sum, aes(vegetative, flowers, color = moisture)) + 
  geom_point() +
  geom_smooth()
```

## Faceted plots

Categorical variables can also be represented by making multiple plots. Add facets to a `ggplot` and specify the variable or variables with `facet_wrap(~varA)` or `facet_grid(varB~varA)`. 

\scriptsize

```{r, fig.width=8, fig.height=3.5, warning=FALSE}
ggplot(SeedPred, aes(tint, taken))+ # data in emdbook
  geom_point()+
  facet_wrap(~species)
```


## Standard routines - numeric and categorical

If the response is numeric and all predictors are categorical, you have some options.

\scriptsize

```{r, fig.width=8, fig.height=3}
g0 <- ggplot(OrchardSprays,aes(x=treatment,y=decrease))+ # data in MASS
  scale_y_log10()
g_boxplot <- g0 + geom_boxplot()
g_point <- g0 +geom_point()
g_errbar <- g0 + stat_summary(fun.data=mean_cl_normal,geom="pointrange")
g_dyn <- g0 +  stat_summary(fun=mean,geom="bar")+
  stat_summary(fun.data=mean_cl_normal,geom="errorbar",width=0.5)
grid.arrange(g_boxplot,g_point,g_errbar,g_dyn, nrow=1)
```

## Dealing with non-standard tasks

Sometimes you need to reshape or summarize your data to plot what you want. To produce Figure 2.1 from @bolker:

\scriptsize
```{r, fig.width=8, fig.height=3.5, message=FALSE, warning=FALSE}
SeedPred$date <- ymd(SeedPred$date) # data in emdbook
daily_avgs <- SeedPred %>%
  group_by(date, species, dist) %>%
  summarise(mean_seeds = mean(seeds))
ggplot(daily_avgs, aes(date, mean_seeds, color=species, shape=dist)) +
  geom_point() + geom_line() + scale_y_log10()
```

## Diagnostics

Assessing the validity of a model is often done graphically.
\scriptsize
```{r, fig.width=8, fig.height=4}
lm1 <- lm(flowers~vegetative, data = Lily_sum)
par(mfrow=c(2, 2), mar = c(4, 4, 2, 2)) # see all 4 plots at once
plot(lm1)
par(mfrow=c(1, 1), mar = c(4, 4, 0.75, 0.5)) # restore graphics parameters
```

## Residuals v. predictors

The `plot` method for `lm` doesn't show residuals against predictors.

\scriptsize
```{r, fig.width=8, fig.height=4}
plot(lm1$residuals~Lily_sum$vegetative)
abline(h=0)
```

## Exposition

Fine tune and save your graphics.
\scriptsize
```{r, fig.width=8, fig.height=3.5, message=FALSE, warning=FALSE}
emd2.1<-ggplot(daily_avgs,aes(date,mean_seeds,color=species,shape=dist))+
  geom_point() + geom_line() + scale_y_log10() +
  labs(y="Mean seeds remaining", x = "Date", 
       color = "Species", shape = "Distance to Forest") +
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
emd2.1
```
```{r, eval =FALSE}
ggsave("figures/BolkerFig2.1.tiff", plot=emd2.1, 
       width = 10, height = 6, units = "cm", dpi = 800)
```

## References

